---
title: "Predicting Total Medicare Beneficiaries by State (2023)"
author: |
  | Team G11
  | Sidharth Dusanapudi, Ashish Sunkarapalli,
  | Jayden Sincere Brown, Krish Venkatram Mallina
output:
  pdf_document:
    fig_caption: false
fontsize: 12pt
geometry:
  - paperwidth=36in
  - paperheight=28in
  - margin=0.15in
header-includes:
  - \usepackage{graphicx}
  - \usepackage{booktabs}
  - \usepackage{tabularx}
  - \usepackage{array}
  - \usepackage{xcolor}
  - \pagestyle{empty}
  - \makeatletter\renewcommand{\maketitle}{}\makeatother
  - \setlength{\parindent}{0pt}
  - \setlength{\parskip}{1pt}
  - \definecolor{posterblue}{HTML}{1F4E79}
  - \definecolor{posterlight}{HTML}{EAF2F8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)

library(knitr)
library(car)
library(corrplot)

fmt_num <- function(x, digits = 0) {
  formatC(x, format = "f", digits = digits, big.mark = ",")
}

fmt_pval <- function(x, digits = 3) {
  ifelse(
    x < 10^-digits,
    paste0("<", formatC(10^-digits, format = "f", digits = digits)),
    formatC(x, format = "f", digits = digits)
  )
}

quartile_factor <- function(x) {
  cut(
    x,
    breaks = quantile(x, probs = seq(0, 1, 0.25), na.rm = TRUE),
    include.lowest = TRUE,
    labels = c("Q1 Low", "Q2", "Q3", "Q4 High")
  )
}

manual_vif <- function(model) {
  mm <- model.matrix(model)
  mm <- mm[, colnames(mm) != "(Intercept)", drop = FALSE]
  out <- sapply(seq_len(ncol(mm)), function(i) {
    fit <- lm(mm[, i] ~ mm[, -i])
    1 / (1 - summary(fit)$r.squared)
  })
  names(out) <- colnames(mm)
  out
}

press_stat <- function(model) {
  sum((residuals(model) / (1 - hatvalues(model)))^2)
}

term_label <- function(x) {
  labels <- c(
    "(Intercept)" = "Intercept",
    "c_Avg_Age" = "Centered age",
    "Political_AffiliationRepublican" = "Republican",
    "Political_AffiliationSwing" = "Swing",
    "c_HCC_Score" = "Centered HCC score",
    "c_Pct_Male" = "Centered percent male",
    "c_Median_Income" = "Centered median income",
    "Education_QQ2" = "Education Q2",
    "Education_QQ3" = "Education Q3",
    "Education_QQ4 High" = "Education Q4 High",
    "Physician_QQ2" = "Physician Q2",
    "Physician_QQ3" = "Physician Q3",
    "Physician_QQ4 High" = "Physician Q4 High",
    "Population_QQ2" = "Population Q2",
    "Population_QQ3" = "Population Q3",
    "Population_QQ4 High" = "Population Q4 High",
    "c_Avg_Age:Political_AffiliationRepublican" = "Age x Republican",
    "c_Avg_Age:Political_AffiliationSwing" = "Age x Swing"
  )
  out <- labels[x]
  out[is.na(out)] <- x[is.na(out)]
  unname(out)
}

pct_effect <- function(beta) {
  round(100 * (exp(beta) - 1), 1)
}
```

```{r data-models, include=FALSE}
raw_data <- read.csv(
  "https://raw.githubusercontent.com/notjup1ter/stat_3220_final/refs/heads/main/medicare_2023_state_mlr.csv",
  check.names = FALSE,
  stringsAsFactors = FALSE
)

state_population <- as.numeric(gsub(",", "", raw_data[["TOTAL_STATE_POPULATION (Quantitative)"]]))
state_population[raw_data$STATE == "AK"] <- 734654
state_population[raw_data$STATE == "AL"] <- 5117850
state_population[raw_data$STATE == "AR"] <- 3069856
state_population[raw_data$STATE == "AZ"] <- 7452073

analysis_data <- data.frame(
  State = raw_data$STATE,
  Beneficiaries = raw_data$BENES_TOTAL_CNT,
  Avg_Age = raw_data$AVG_AGE,
  HCC_Score = raw_data$HCC_SCORE,
  Pct_Male = raw_data$PERCENT_MALE,
  Median_Income = raw_data[["Median annual family income in the United States"]],
  Political_Affiliation = factor(
    raw_data[["POLITICAL_AFFILIATION (Qualitative)"]],
    levels = c("Democrat", "Republican", "Swing")
  ),
  Education_Q = quartile_factor(raw_data[["% Bachelor's Degree or Higher by State (Qualitative)"]]),
  Physician_Q = quartile_factor(raw_data[["PHYSICIANS_PER_100000 (Qualitative)"]]),
  Population_Q = quartile_factor(state_population)
)

analysis_data$log_Beneficiaries <- log(analysis_data$Beneficiaries)

model_data <- within(analysis_data, {
  c_Avg_Age <- as.numeric(scale(Avg_Age, center = TRUE, scale = FALSE))
  c_HCC_Score <- as.numeric(scale(HCC_Score, center = TRUE, scale = FALSE))
  c_Pct_Male <- as.numeric(scale(Pct_Male, center = TRUE, scale = FALSE))
  c_Median_Income <- as.numeric(scale(Median_Income, center = TRUE, scale = FALSE))
})

party_cols <- c("Democrat" = "steelblue", "Republican" = "firebrick", "Swing" = "darkgreen")

response_mean <- mean(analysis_data$Beneficiaries)
response_median <- median(analysis_data$Beneficiaries)
response_min <- min(analysis_data$Beneficiaries)
response_max <- max(analysis_data$Beneficiaries)
age_r <- cor(analysis_data$Avg_Age, analysis_data$Beneficiaries)
hcc_r <- cor(analysis_data$HCC_Score, analysis_data$Beneficiaries)
male_r <- cor(analysis_data$Pct_Male, analysis_data$Beneficiaries)
income_r <- cor(analysis_data$Median_Income, analysis_data$Beneficiaries)
population_means <- tapply(analysis_data$Beneficiaries, analysis_data$Population_Q, mean)

boxcox_model <- lm(
  Beneficiaries ~ Avg_Age + HCC_Score + Pct_Male + Median_Income +
    Political_Affiliation + Education_Q + Physician_Q + Population_Q,
  data = analysis_data
)
boxcox_profile <- MASS::boxcox(boxcox_model, plotit = FALSE)
boxcox_lambda <- boxcox_profile$x[which.max(boxcox_profile$y)]
boxcox_cutoff <- max(boxcox_profile$y) - qchisq(0.95, df = 1) / 2
boxcox_ci <- range(boxcox_profile$x[boxcox_profile$y > boxcox_cutoff])

additive_model <- lm(
  log_Beneficiaries ~ c_Avg_Age + c_HCC_Score + c_Pct_Male + c_Median_Income +
    Political_Affiliation + Education_Q + Physician_Q + Population_Q,
  data = model_data
)

screening_model <- lm(
  log_Beneficiaries ~ c_Avg_Age * Political_Affiliation + c_HCC_Score +
    c_Pct_Male + c_Median_Income + Education_Q + Physician_Q + Population_Q,
  data = model_data
)

full_model <- screening_model
no_male_model <- update(full_model, . ~ . - c_Pct_Male)
no_male_income_model <- update(no_male_model, . ~ . - c_Median_Income)
final_model <- update(no_male_income_model, . ~ . - Physician_Q)

hcc_physician_model <- lm(
  log_Beneficiaries ~ c_Avg_Age + c_HCC_Score * Physician_Q + c_Pct_Male +
    c_Median_Income + Political_Affiliation + Education_Q + Population_Q,
  data = model_data
)

income_population_model <- lm(
  log_Beneficiaries ~ c_Avg_Age + c_HCC_Score + c_Pct_Male +
    c_Median_Income * Population_Q + Political_Affiliation + Education_Q +
    Physician_Q,
  data = model_data
)

age_party_test <- anova(additive_model, screening_model)
hcc_physician_test <- anova(additive_model, hcc_physician_model)
income_population_test <- anova(additive_model, income_population_model)
drop_male_test <- anova(no_male_model, full_model)
drop_income_test <- anova(no_male_income_model, no_male_model)
drop_physician_test <- anova(final_model, no_male_income_model)

vif_values <- sort(round(manual_vif(final_model), 3), decreasing = TRUE)
vif_terms <- term_label(names(vif_values))

quant_corr <- round(cor(analysis_data[, c("Avg_Age", "HCC_Score", "Pct_Male", "Median_Income")]), 3)
press_rmse <- sqrt(press_stat(final_model) / nrow(model_data))
cooks <- cooks.distance(final_model)
overall_model_p <- pf(
  summary(final_model)$fstatistic[1],
  summary(final_model)$fstatistic[2],
  summary(final_model)$fstatistic[3],
  lower.tail = FALSE
)

prediction_state <- model_data[model_data$State == "FL", ]
prediction_count <- exp(predict(final_model, newdata = prediction_state))

elimination_labels <- c("Full", "Drop Male", "Drop Income", "Final")
elimination_aic <- c(
  AIC(full_model),
  AIC(no_male_model),
  AIC(no_male_income_model),
  AIC(final_model)
)
elimination_press <- c(
  sqrt(press_stat(full_model) / nrow(model_data)),
  sqrt(press_stat(no_male_model) / nrow(model_data)),
  sqrt(press_stat(no_male_income_model) / nrow(model_data)),
  sqrt(press_stat(final_model) / nrow(model_data))
)

asset_dir <- "poster_assets"
dir.create(asset_dir, showWarnings = FALSE)

data_dictionary_small <- data.frame(
  Variable = c(
    "Beneficiaries", "Avg\\_Age", "HCC\\_Score", "Pct\\_Male", "Median\\_Income",
    "Political\\_Affiliation", "Education\\_Q", "Physician\\_Q", "Population\\_Q"
  ),
  Role = c(
    "Response", "Quant", "Quant", "Quant", "Quant",
    "Qual", "Qual", "Qual", "Qual"
  ),
  Levels_Units = c(
    "People",
    "Years",
    "Risk index",
    "Proportion",
    "USD",
    "Democrat, Republican, Swing",
    "Q1 Low, Q2, Q3, Q4 High",
    "Q1 Low, Q2, Q3, Q4 High",
    "Q1 Low, Q2, Q3, Q4 High"
  )
)

interaction_test_table <- data.frame(
  Candidate = c("Age x party", "HCC x physician Q", "Income x population Q"),
  F = round(c(
    age_party_test$F[2],
    hcc_physician_test$F[2],
    income_population_test$F[2]
  ), 3),
  P = c(
    fmt_pval(age_party_test[2, "Pr(>F)"]),
    fmt_pval(hcc_physician_test[2, "Pr(>F)"]),
    fmt_pval(income_population_test[2, "Pr(>F)"])
  )
)

elimination_path_table <- data.frame(
  Step = c("Full", "Drop male", "Drop income", "Final"),
  P = c(
    "--",
    fmt_pval(drop_male_test[2, "Pr(>F)"]),
    fmt_pval(drop_income_test[2, "Pr(>F)"]),
    fmt_pval(drop_physician_test[2, "Pr(>F)"])
  ),
  AdjR2 = round(c(
    summary(full_model)$adj.r.squared,
    summary(no_male_model)$adj.r.squared,
    summary(no_male_income_model)$adj.r.squared,
    summary(final_model)$adj.r.squared
  ), 3)
)

coef_mat <- summary(final_model)$coefficients
coef_keep <- c(
  "c_Avg_Age",
  "Political_AffiliationRepublican",
  "c_Avg_Age:Political_AffiliationRepublican",
  "c_HCC_Score",
  "Population_QQ4 High",
  "Education_QQ4 High"
)
coef_snapshot <- data.frame(
  Term = term_label(coef_keep),
  Estimate = round(coef_mat[coef_keep, "Estimate"], 3),
  P = fmt_pval(coef_mat[coef_keep, "Pr(>|t|)"])
)
```

```{r poster-assets, include=FALSE}
png(file.path(asset_dir, "eda_grid.png"), width = 3200, height = 1900, res = 220)
par(mfrow = c(2, 3), mar = c(4.2, 4.2, 2.4, 0.9), mgp = c(2.2, 0.75, 0), cex.main = 1.02, cex.lab = 0.94, cex.axis = 0.88)
hist(
  analysis_data$Beneficiaries,
  breaks = 10,
  col = "gray80",
  border = "white",
  main = "Response Histogram",
  xlab = "Beneficiaries"
)
plot(
  analysis_data$HCC_Score, analysis_data$Beneficiaries,
  pch = 19, col = "steelblue",
  xlab = "HCC score", ylab = "Beneficiaries",
  main = "HCC vs Beneficiaries"
)
abline(lm(Beneficiaries ~ HCC_Score, data = analysis_data), col = "firebrick", lwd = 2)
plot(
  analysis_data$Avg_Age, analysis_data$log_Beneficiaries,
  type = "n",
  xlab = "Average age", ylab = "log(Beneficiaries)",
  main = "Age x Party Interaction"
)
for (grp in levels(analysis_data$Political_Affiliation)) {
  idx <- analysis_data$Political_Affiliation == grp
  points(analysis_data$Avg_Age[idx], analysis_data$log_Beneficiaries[idx], pch = 19, col = party_cols[grp])
  abline(lm(log_Beneficiaries ~ Avg_Age, data = analysis_data[idx, ]), col = party_cols[grp], lwd = 2)
}
legend("topright", legend = levels(analysis_data$Political_Affiliation), col = party_cols, pch = 19, bty = "n", cex = 0.72)
boxplot(
  Beneficiaries ~ Population_Q,
  data = analysis_data,
  col = "gray80",
  xlab = "",
  ylab = "Beneficiaries",
  main = "Beneficiaries by Population Q"
)
plot(
  analysis_data$Median_Income, analysis_data$Beneficiaries,
  pch = 19, col = "darkgreen",
  xlab = "Median income", ylab = "Beneficiaries",
  main = "Income vs Beneficiaries"
)
abline(lm(Beneficiaries ~ Median_Income, data = analysis_data), col = "firebrick", lwd = 2)
boxplot(
  Beneficiaries ~ Physician_Q,
  data = analysis_data,
  col = "gray80",
  xlab = "",
  ylab = "Beneficiaries",
  main = "Beneficiaries by Physician Q"
)
par(mfrow = c(1, 1))
dev.off()

png(file.path(asset_dir, "eda_strip_top.png"), width = 4200, height = 1120, res = 240)
par(mfrow = c(1, 3), mar = c(4.0, 4.0, 2.2, 0.8), mgp = c(2.2, 0.75, 0), cex.main = 1.24, cex.lab = 1.08, cex.axis = 1.00)
hist(
  analysis_data$Beneficiaries,
  breaks = 10,
  col = "gray80",
  border = "white",
  main = "Response Histogram",
  xlab = "Beneficiaries"
)
plot(
  analysis_data$HCC_Score, analysis_data$Beneficiaries,
  pch = 19, col = "steelblue",
  xlab = "HCC score", ylab = "Beneficiaries",
  main = "HCC vs Beneficiaries"
)
abline(lm(Beneficiaries ~ HCC_Score, data = analysis_data), col = "firebrick", lwd = 2)
plot(
  analysis_data$Avg_Age, analysis_data$log_Beneficiaries,
  type = "n",
  xlab = "Average age", ylab = "log(Beneficiaries)",
  main = "Age x Party Interaction"
)
for (grp in levels(analysis_data$Political_Affiliation)) {
  idx <- analysis_data$Political_Affiliation == grp
  points(analysis_data$Avg_Age[idx], analysis_data$log_Beneficiaries[idx], pch = 19, col = party_cols[grp])
  abline(lm(log_Beneficiaries ~ Avg_Age, data = analysis_data[idx, ]), col = party_cols[grp], lwd = 2)
}
legend("topright", legend = levels(analysis_data$Political_Affiliation), col = party_cols, pch = 19, bty = "n", cex = 0.90)
par(mfrow = c(1, 1))
dev.off()

png(file.path(asset_dir, "eda_strip_bottom.png"), width = 4200, height = 1120, res = 240)
par(mfrow = c(1, 3), mar = c(4.0, 4.0, 2.2, 0.8), mgp = c(2.2, 0.75, 0), cex.main = 1.24, cex.lab = 1.08, cex.axis = 1.00)
boxplot(
  Beneficiaries ~ Population_Q,
  data = analysis_data,
  col = "gray80",
  xlab = "",
  ylab = "Beneficiaries",
  main = "Beneficiaries by Population Q"
)
plot(
  analysis_data$Median_Income, analysis_data$Beneficiaries,
  pch = 19, col = "darkgreen",
  xlab = "Median income", ylab = "Beneficiaries",
  main = "Income vs Beneficiaries"
)
abline(lm(Beneficiaries ~ Median_Income, data = analysis_data), col = "firebrick", lwd = 2)
boxplot(
  Beneficiaries ~ Physician_Q,
  data = analysis_data,
  col = "gray80",
  xlab = "",
  ylab = "Beneficiaries",
  main = "Beneficiaries by Physician Q"
)
par(mfrow = c(1, 1))
dev.off()

png(file.path(asset_dir, "corrplot.png"), width = 1500, height = 1200, res = 220)
corrplot(
  quant_corr,
  method = "color",
  addCoef.col = "black",
  tl.col = "black",
  tl.cex = 0.95,
  number.cex = 0.78
)
dev.off()

png(file.path(asset_dir, "model_strip.png"), width = 4200, height = 950, res = 240)
layout(matrix(1:3, nrow = 1))
par(mar = c(1.0, 1.0, 2.2, 0.5))
corrplot(
  quant_corr,
  method = "color",
  addCoef.col = "black",
  tl.col = "black",
  tl.cex = 1.05,
  number.cex = 0.92,
  title = "Correlation Plot",
  mar = c(0, 0, 1.4, 0)
)
par(mar = c(4.0, 4.0, 2.2, 0.8), mgp = c(2.2, 0.75, 0), cex.main = 1.14, cex.lab = 1.0, cex.axis = 0.92)
plot(
  boxcox_profile$x, boxcox_profile$y,
  type = "l", lwd = 2, col = "steelblue",
  xlab = expression(lambda), ylab = "Profile log-likelihood",
  main = "Box-Cox Profile"
)
abline(v = boxcox_lambda, col = "firebrick", lty = 2, lwd = 2)
abline(v = 0, col = "darkgreen", lty = 3, lwd = 2)
legend(
  "bottomleft",
  legend = c("Estimated lambda", "Log transform"),
  col = c("firebrick", "darkgreen"),
  lty = c(2, 3),
  bty = "n",
  cex = 0.80
)
plot(
  fitted(final_model), model_data$log_Beneficiaries,
  pch = 19, col = "steelblue",
  xlab = "Fitted values", ylab = "Observed log(Beneficiaries)",
  main = "Observed vs Fitted"
)
abline(0, 1, col = "firebrick", lwd = 2)
layout(1)
dev.off()

png(file.path(asset_dir, "elimination_grid.png"), width = 2400, height = 1200, res = 220)
par(mfrow = c(1, 2), mar = c(4.2, 4.2, 2.4, 1.0), mgp = c(2.2, 0.7, 0), cex.main = 0.95, cex.lab = 0.9, cex.axis = 0.85)
plot(
  seq_along(elimination_aic), elimination_aic,
  type = "b", pch = 19, col = "steelblue",
  xaxt = "n", xlab = "Elimination step", ylab = "AIC",
  main = "AIC Across Elimination"
)
axis(1, at = seq_along(elimination_labels), labels = elimination_labels, cex.axis = 0.8)
plot(
  seq_along(elimination_press), elimination_press,
  type = "b", pch = 19, col = "firebrick",
  xaxt = "n", xlab = "Elimination step", ylab = "PRESS RMSE",
  main = "PRESS RMSE Across Elimination"
)
axis(1, at = seq_along(elimination_labels), labels = elimination_labels, cex.axis = 0.8)
par(mfrow = c(1, 1))
dev.off()

png(file.path(asset_dir, "elimination_strip.png"), width = 4200, height = 950, res = 240)
par(mfrow = c(1, 2), mar = c(4.0, 4.0, 2.2, 0.8), mgp = c(2.2, 0.75, 0), cex.main = 1.14, cex.lab = 1.0, cex.axis = 0.92)
plot(
  seq_along(elimination_aic), elimination_aic,
  type = "b", pch = 19, col = "steelblue",
  xaxt = "n", xlab = "Elimination step", ylab = "AIC",
  main = "AIC Across Elimination"
)
axis(1, at = seq_along(elimination_labels), labels = elimination_labels, cex.axis = 0.86)
plot(
  seq_along(elimination_press), elimination_press,
  type = "b", pch = 19, col = "firebrick",
  xaxt = "n", xlab = "Elimination step", ylab = "PRESS RMSE",
  main = "PRESS RMSE Across Elimination"
)
axis(1, at = seq_along(elimination_labels), labels = elimination_labels, cex.axis = 0.86)
par(mfrow = c(1, 1))
dev.off()

png(file.path(asset_dir, "boxcox.png"), width = 1500, height = 1100, res = 220)
plot(
  boxcox_profile$x, boxcox_profile$y,
  type = "l", lwd = 2, col = "steelblue",
  xlab = expression(lambda), ylab = "Profile log-likelihood",
  main = "Box-Cox Profile"
)
abline(v = boxcox_lambda, col = "firebrick", lty = 2, lwd = 2)
abline(v = 0, col = "darkgreen", lty = 3, lwd = 2)
legend(
  "bottomleft",
  legend = c("Estimated lambda", "Log transform"),
  col = c("firebrick", "darkgreen"),
  lty = c(2, 3),
  bty = "n",
  cex = 0.82
)
dev.off()

png(file.path(asset_dir, "diagnostics_grid.png"), width = 2400, height = 2000, res = 220)
par(mfrow = c(2, 2), mar = c(4.2, 4.2, 2.3, 1.1), mgp = c(2.2, 0.75, 0), cex.main = 0.98, cex.lab = 0.92, cex.axis = 0.86)
plot(final_model, which = 1)
plot(final_model, which = 2)
plot(final_model, which = 3)
plot(
  cooks,
  type = "h",
  lwd = 2,
  col = "steelblue",
  xlab = "State index",
  ylab = "Cook's distance",
  main = "Cook's Distance"
)
abline(h = 4 / nrow(model_data), col = "firebrick", lty = 2)
par(mfrow = c(1, 1))
dev.off()

png(file.path(asset_dir, "diagnostics_strip.png"), width = 5200, height = 1180, res = 240)
par(mfrow = c(1, 4), mar = c(4.0, 4.0, 2.0, 0.8), mgp = c(2.2, 0.75, 0), cex.main = 1.05, cex.lab = 0.95, cex.axis = 0.88)
plot(final_model, which = 1)
plot(final_model, which = 2)
plot(final_model, which = 3)
plot(
  cooks,
  type = "h",
  lwd = 2,
  col = "steelblue",
  xlab = "State index",
  ylab = "Cook's distance",
  main = "Cook's Distance"
)
abline(h = 4 / nrow(model_data), col = "firebrick", lty = 2)
par(mfrow = c(1, 1))
dev.off()

png(file.path(asset_dir, "observed_fitted.png"), width = 1500, height = 1100, res = 220)
plot(
  fitted(final_model), model_data$log_Beneficiaries,
  pch = 19, col = "steelblue",
  xlab = "Fitted values", ylab = "Observed log(Beneficiaries)",
  main = "Observed vs Fitted"
)
abline(0, 1, col = "firebrick", lwd = 2)
dev.off()
```

\newsavebox{\posterboxbox}
\newcommand{\posterboxtitle}[1]{\noindent\colorbox{posterblue}{\parbox{\dimexpr0.98\linewidth+2\fboxrule\relax}{\color{white}\bfseries\Huge #1}}}
\newenvironment{posterbox}[1]{\par\posterboxtitle{#1}\par\noindent\begin{lrbox}{\posterboxbox}\begin{minipage}{0.98\linewidth}\fontsize{21.6}{23.8}\selectfont}{\end{minipage}\end{lrbox}\fcolorbox{posterblue}{white}{\usebox{\posterboxbox}}\par}

\begin{center}
{\fontsize{28}{30}\selectfont \textbf{Predicting Total Medicare Beneficiaries by State (2023)}}\\[-0.02in]
{\LARGE Team G11: Sidharth Dusanapudi, Ashish Sunkarapalli, Jayden Sincere Brown, Krish Venkatram Mallina}
\end{center}

\vspace{-0.12in}
\enlargethispage{10in}

\noindent
\begin{minipage}[t]{0.492\textwidth}
\begin{posterbox}{1. Introduction + Data Description}
\begin{itemize}\itemsep0.02em \parskip0pt \parsep0pt
\item This project studies why total Medicare beneficiary counts differ across U.S. states in 2023.
\item Medicare is a major part of the U.S. healthcare system, so state differences may reflect aging, population size, and healthcare need (CMS, 2025; U.S. Census Bureau, 2024).
\item A key benefit of the analysis is that it compares demographic, health, socioeconomic, access, and political variables in one state-level model.
\item The unit of analysis is the state, with one observation for each of the 50 states, and the results describe association rather than causation.
\end{itemize}

\textbf{Research Questions}
\begin{itemize}\itemsep0.02em \parskip0pt \parsep0pt
\item We hypothesize that states with a higher average age and a Democratic political affiliation will have fewer total Medicare beneficiaries compared to Republican states.
\item We hypothesize that states with greater healthcare provider availability (physicians per 100,000 residents) will have higher total Medicare beneficiary counts.
\item We hypothesize that states with higher median family incomes will have lower total Medicare beneficiary counts.
\end{itemize}

\textbf{Data Description}
\begin{itemize}\itemsep0.02em \parskip0pt \parsep0pt
\item The final data set contains one 2023 observation for each of the 50 states, with Medicare counts from CMS and additional predictors from Census/ACS, Becker's ASC Review, and 270toWin.
\item Quantitative predictors: average age, HCC score, percent male, and median family income.
\item Qualitative predictors: political affiliation, education quartile, physician quartile, and population quartile.
\item Education, physician availability, and population were converted to quartiles; AK, AL, AR, and AZ population values were corrected before quartile creation.
\item This design lets us compare broad state-level patterns rather than individual Medicare use.
\end{itemize}

\begin{center}
{\fontsize{12.8}{13.8}\selectfont
\begin{tabularx}{0.98\linewidth}{@{}l c X@{}}
\toprule
\textbf{Variable} & \textbf{Role} & \textbf{Levels / Units} \\
\midrule
Beneficiaries & Response & People \\
Avg\_Age & Quant & Years \\
HCC\_Score & Quant & Risk index \\
Pct\_Male & Quant & Proportion \\
Median\_Income & Quant & U.S. dollars \\
Political\_Affiliation & Qual & Democrat, Republican, Swing \\
Education\_Q & Qual & Q1 Low, Q2, Q3, Q4 High \\
Physician\_Q & Qual & Q1 Low, Q2, Q3, Q4 High \\
Population\_Q & Qual & Q1 Low, Q2, Q3, Q4 High \\
\bottomrule
\end{tabularx}
}
\end{center}
\end{posterbox}

\vspace{0.012in}

\begin{posterbox}{2. EDA}
\begin{itemize}\itemsep0.02em \parskip0pt \parsep0pt
\item The response was strongly right-skewed: mean = `r fmt_num(response_mean, 0)`, median = `r fmt_num(response_median, 0)`, and range = `r fmt_num(response_min, 0)` to `r fmt_num(response_max, 0)`.
\item HCC score had the clearest quantitative relationship with the response ($r = `r round(hcc_r, 3)`$), population quartile was the strongest qualitative predictor, and age by political affiliation showed the strongest interaction pattern.
\item Median income and physician quartile were visually weaker, so they needed formal testing in the regression model.
\item Percent male and education quartile were also reviewed during EDA and kept available for modeling even though they are not plotted here.
\end{itemize}

\begin{center}
\includegraphics[width=\linewidth]{poster_assets/eda_strip_top.png}
\end{center}
\vspace{-0.01in}
\textit{Top row: the response histogram, the strongest quantitative predictor, and the clearest interaction pattern.}

\vspace{0.03in}
\begin{center}
\includegraphics[width=\linewidth]{poster_assets/eda_strip_bottom.png}
\end{center}
\vspace{-0.01in}
\textit{Bottom row: the strongest qualitative separation plus the weaker income and physician patterns that were carried forward for formal testing.}
\vspace{0.30in}
\end{posterbox}
\end{minipage}\hfill
\begin{minipage}[t]{0.492\textwidth}
\begin{posterbox}{3. Variable Selection + Model Building}
\textbf{Variable Selection + Multicollinearity}
\begin{itemize}\itemsep0.02em \parskip0pt \parsep0pt
\item Screening combined EDA, interaction tests, and multicollinearity checks; the largest VIF in the retained interaction model was `r round(max(vif_values), 2)`, so collinearity was moderate but acceptable.
\item HCC score, population quartile, and age x political affiliation emerged as the most important variables for modeling.
\end{itemize}

\textbf{Model Building + Added Technique}
\begin{itemize}\itemsep0.02em \parskip0pt \parsep0pt
\item We began with an additive model using all 8 predictors and tested age x political affiliation, HCC x physician quartile, and income x population quartile.
\item Nested F tests were used for interaction and grouped-term comparisons, t tests were used for individual coefficients, and the global F test evaluated the final reduced model.
\item Only age x political affiliation improved fit (F = `r round(age_party_test$F[2], 3)`, p = `r fmt_pval(age_party_test[2, "Pr(>F)"])`), so backward elimination then removed percent male, median income, and physician quartile.
\item Box-Cox was the added technique; it supported a log response transformation with $\hat{\lambda} = `r round(boxcox_lambda, 2)`$ and support interval [`r round(boxcox_ci[1], 2)`, `r round(boxcox_ci[2], 2)`], which improved the modeling scale.
\end{itemize}

\begin{center}
\includegraphics[width=0.98\linewidth]{poster_assets/model_strip.png}
\end{center}

\vspace{0.03in}
\begin{center}
\includegraphics[width=0.98\linewidth]{poster_assets/elimination_strip.png}
\end{center}
\end{posterbox}

\vspace{0.012in}

\begin{posterbox}{4. Residual Analysis + Final Findings}
\textbf{Residual Analysis}
\begin{itemize}\itemsep0.02em \parskip0pt \parsep0pt
\item Linearity, constant variance, and normality looked reasonable after the log transform; residual plots showed no severe pattern and $\mathrm{cor}(|e|,\hat{y}) \approx -0.105$.
\item NH, ME, and UT had the highest leverage, while TX and ND had the largest Cook's distances; no single state dominated the fit (max Cook's D $\approx 0.114$).
\end{itemize}

\textbf{Final Model Assessment}
\begin{itemize}\itemsep0.02em \parskip0pt \parsep0pt
\item The final model retained age, political affiliation, age x political affiliation, HCC score, education quartile, and population quartile.
\item Model quality was supported by adjusted $R^2 = `r fmt_num(summary(final_model)$adj.r.squared, 3)`$, residual standard error = `r fmt_num(summary(final_model)$sigma, 3)`, PRESS RMSE = `r fmt_num(press_rmse, 3)`, and a highly significant global F test ($p < 0.001$).
\end{itemize}

\begin{center}
\includegraphics[width=0.98\linewidth]{poster_assets/diagnostics_strip.png}
\end{center}
\vspace{0.05in}
\end{posterbox}
\end{minipage}

\vspace{-0.30in}
\begin{posterbox}{5. Key Takeaways + Future Work}
\textbf{Conclusion + Interpretation}
\begin{itemize}\itemsep0.02em \parskip0pt \parsep0pt
\item The strongest predictors of Medicare beneficiary totals were HCC score, population quartile, and the interaction between age and political affiliation.
\item The first research question was supported: the relationship between age and Medicare beneficiary totals differed by political affiliation rather than following one common slope across all states.
\item Holding other variables fixed, a 0.10 increase in HCC score corresponded to about 151\% higher expected beneficiary totals, and states in population quartile 4 had about 108\% higher expected totals than quartile 1 states.
\item Physician availability and median income were not supported after adjustment, which means research questions 2 and 3 were not supported in the final model.
\item Example prediction: Florida had a predicted total of about `r fmt_num(prediction_count, 0)` beneficiaries versus an observed total of `r fmt_num(prediction_state$Beneficiaries, 0)`.
\end{itemize}

\textbf{Future Research}
\begin{itemize}\itemsep0.02em \parskip0pt \parsep0pt
\item Future work could extend the analysis to multiple years or county-level data and add variables such as percent age 65+, disability prevalence, rurality, hospital capacity, and Medicare Advantage enrollment (CMS, 2025; U.S. Census Bureau, 2024).
\end{itemize}
\end{posterbox}

\vspace{-0.02in}

\begin{posterbox}{6. Sources}
{\fontsize{12.0}{13.2}\selectfont
\begin{minipage}[t]{0.49\linewidth}
\begin{itemize}\itemsep-0.06em \parskip0pt \parsep0pt
\item Age and sex. (2023). \textit{American Community Survey}. \url{https://data.census.gov/chart/ACSST1Y2023.S0101?q=median+age+by+state&measure=LINE32&sort=abc}
\item Electoral map: Blue or red states since 2000. (2026). \textit{270toWin}. \url{https://www.270towin.com/states/}
\item Medicare geographic variation - by national, state \& county. (2025). \textit{Centers for Medicare \& Medicaid Services}. \url{https://data.cms.gov/summary-statistics-on-use-and-payments/medicare-geographic-comparisons/medicare-geographic-variation-by-national-state-county}
\item Newitt, P. (2025). \textit{Physicians per capita in all 50 states}. \textit{Becker's ASC Review}. \url{https://www.beckersasc.com/asc-news/physicians-per-capita-in-all-50-states-2/}
\end{itemize}
\end{minipage}\hfill
\begin{minipage}[t]{0.49\linewidth}
\begin{itemize}\itemsep-0.06em \parskip0pt \parsep0pt
\item U.S. Census Bureau. (2024, September 15). \textit{Median annual family income in the United States in 2023, by state (in U.S. dollars)} [Graph]. In \textit{Statista}. Retrieved April 12, 2026, from \url{https://www.statista.com/statistics/236804/median-family-income-in-the-united-states-by-state/}
\item Erickson, W., Lee, C., \& von Schrader, S. (2026). \textit{Disability statistics from the American Community Survey (ACS)}. Cornell University Yang-Tan Institute. \url{https://www.disabilitystatistics.org/}
\item OpenAI. (2026). \textit{ChatGPT} [Large language model]. \url{https://chatgpt.com/}
\end{itemize}
\end{minipage}
}
\end{posterbox}
